Spatial Reasoning and Data Displays

### 2015-01-28 Paper/Draft1/TestingVisualAptitude.Rnw
Remove extraneous comment
still figure 6

\newtext{Principal component analysis separates multivariate data into orthogonal components using a rotation matrix, which contains information on the contribution of each input variable to the principal component. The importance of a principal component is also evaluated to assess the proportion of the overall variance in the data contained within the component (table \ref{tab:PCAvariance5} shows the importance of each PC for the analysis of the aggregate lineup score and the four aptitude tests).}

<<lineupblockpca, echo=F, warning=F, message=F, fig.cap="Principal Component importance plot, showing each input variable's contribution to the principal component, scaled by the proportion of variability in the data contained in the principal component.", fig.width=5, fig.height=5>>=
lineup.block.pca <- prcomp(lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3", "card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
rotation <- melt(lineup.block.pca$rotation)
importance <- as.data.frame(summary(lineup.block.pca)$importance)[2,]
rotation <- merge(rotation, melt(importance, variable.name="Var2", value.name="importance"))
rotation$Var1 <- factor(rotation$Var1, 
                        levels=c("vis_search", "folding", "card_rot", "fig_class", "lineup_3", "lineup_2", "lineup_1"),
                        labels=c("Visual Search", "Paper Folding", "Card Rotation", "Figure Class.", "Lineup Test 3", "Lineup Test 2", "Lineup Test 1"))
library(grid) # needed for arrow function
qplot(x=0,  y=Var1, xend=value*importance, yend=Var1, geom="segment", data=rotation, arrow=arrow(length=unit(0.1, "cm"))) + facet_wrap(~Var2, ncol=3) + xlab("Influence") + ggtitle("Lineup Types and Visual Aptitude") + ylab("") + theme_bw()
@

@@

This scaling gets slightly more complicated for scores which have two sub-groups, as with the figure classification test, which separately sumarizes male and female participants' scores. 
To get a single unified score with standard deviation, we completed the following calculations: 
\begin{align}
\mu_{\text{all}} &= (N_F\mu_{F} + N_M\mu_{M})/(N_F + N_M)\\
\sigma_{\text{all}} &= \sqrt{\left(N_F\sigma_F^2 + N_M\sigma_M^2\right)/(N_F + N_M)},
\end{align}
where $\mu_F$ and $\mu_M$ are the mean of female and male scores respectively, $N_F$ and $N_M$ are the number of participants in each group, and $\sigma_F^2$ and $\sigma_M^2$ are the variance of each group.
Substituting in the provided numbers, we get
\begin{align*}
\mu_{\text{all}} &= \left(323\!\cdot\!114.9\! +\! 294\!\cdot\!120.0\right)/(323\!+\!294) \\
& = \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)}\\
\sigma_{\text{all}} &= \sqrt{\left(323(27.8)^2 \!+\! 294(30)^2\right)/(323\!+\!294)} \\
& = \Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)}.
\end{align*}

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.7\\linewidth', comment="|", fig.cap="Biplots of principal components 1-4 with observations. Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca2, choices=1:2, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca2, choices=3:4, pc.biplot=T, cex=c(.5, 1), xlim=c(-3,3), ylim=c(-2.5,2.5))
@
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the following principal components separate between the cognitive tasks.
\comment{How do the PCs exactly separate? Quick sentence for each.}

According to Table~\ref{tab:PCArotation4}, the first three principal components account for \Sexpr{round(summary(pca2)$importance[3,3]*100,1)}\% of the variance.

@@

\comment{Describe performance - need to get data from original experiment.}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}
\caption{Boxplots used to compare the two distributions.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}
\caption{Density plots used to compare the two distributions.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}
\caption{Dotplots used to compare the two distributions.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}
\caption{Histograms used to compare the two distributions.}
\end{figure}
\FloatBarrier
\subsection{Lineup Set 2}
The second lineup section also explored two groups of data, this time comparing boxplots, bee swarm boxplots, boxplots with overlaid jittered data, and violin plots. Participants were much more accurate in this experiment than in the experiment described previously, because of the types of plots compared as well as the underlying data distributions. 
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f94f6965d7}
\caption{Boxplots.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f9e65f0e4}
\caption{Boxplots with jittered points.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f9fc58d73}
\caption{Bee swarm boxplots.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f93acbb7e4}
\caption{Violin plots.}
\end{figure}
\FloatBarrier
\subsection{Lineup Set 3}
The final lineup section explored QQ-plots from various model simulations, using reference lines, acceptance bands, and rotation to determine which plots allowed participants to most effectively identify violations of normality. Rotated qq plots showed lower performance because participants could more accurately compare acceptance bands to residuals, and thus could identify that the reference bands were too liberal. As a result, performance was somewhat lower for rotated plots, even though participants could more accurately compare the residuals to the reference bands.
% \comment{Add citations, more explanations, etc.}

>>>>

\comment{Susan, sorry, the description of figure 6 is still not working. You write impact at the x axis - so I'm wondering what a negative impact is supposed to be? You are mentioning contribution and importance in the paragraph below, but you don't really define either one of them.
Be careful about mentioning Table 2 - that table doesn't have anything to do with this PCA - it might be better to include the two tables for this PCA in the appendix and work out there in detail, how the scaling is done. InfoVis does know about signular value composition, so you could start there ... }

\newtext{
Principal component analysis separates multivariate data into orthogonal components using a rotation matrix, which contains information on the contribution of each input variable to the principal component. The importance of a principal component is also evaluated to assess the proportion of the overall variance in the data contained within the component (Table \ref{tab:PCAvariance5} shows the importance of each PC for the analysis of the aggregate lineup score and the four aptitude tests). 
}
<<lineupblockpca, echo=F, warning=F, message=F, fig.cap="Principal Component importance plot, showing each input variable's contribution to the principal component, scaled by the proportion of variability in the data contained in the principal component.", fig.width=5, fig.height=5>>=
lineup.block.pca <- prcomp(lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3", "card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
rotation <- melt(lineup.block.pca$rotation)
importance <- as.data.frame(summary(lineup.block.pca)$importance)[2,]
rotation <- merge(rotation, melt(importance, variable.name="Var2", value.name="importance"))
rotation$Var1 <- factor(rotation$Var1, 
                        levels=c("vis_search", "folding", "card_rot", "fig_class", "lineup_3", "lineup_2", "lineup_1"),
                        labels=c("Visual Search", "Paper Folding", "Card Rotation", "Figure Class.", "Lineup Test 3", "Lineup Test 2", "Lineup Test 1"))
library(grid) # needed for arrow function
qplot(x=0,  y=Var1, xend=value*importance, yend=Var1, geom="segment", data=rotation, arrow=arrow(length=unit(0.1, "cm"))) + facet_wrap(~Var2, ncol=3) + xlab("Influence") + ggtitle("Lineup Types and Visual Aptitude") + ylab("") + theme_bw()
@

@@

This scaling gets slightly more complicated for scores which have two sub-groups, as with the figure classification test, which separately sumarizes male and female participants' scores. 
To get a single unified score with standard deviation, we completed the following calculations: 
\begin{align}
\mu_{\text{all}} &= (N_F\mu_{F} + N_M\mu_{M})/(N_F + N_M)\\
\sigma_{\text{all}} &= \sqrt{\left(N_F\sigma_F^2 + N_M\sigma_M^2\right)/(N_F + N_M)},
\end{align}
where $\mu_F$ and $\mu_M$ are the mean of female and male scores respectively, $N_F$ and $N_M$ are the number of participants in each group, and $\sigma_F^2$ and $\sigma_M^2$ are the variance of each group.
Substituting in the provided numbers, we get
\begin{align*}
\mu_{\text{all}} &= \left(323\!\cdot\!114.9\! +\! 294\!\cdot\!120.0\right)/(323\!+\!294) \\
& = \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)}\\
\sigma_{\text{all}} &= \sqrt{\left(323 \cdot 27.8^2 \!+\! 294 \cdot 30^2\right)/(323\!+\!294)} \\
& = \Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)}.
\end{align*}

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.7\\linewidth', comment="|", fig.cap="Biplots of principal components 1-4 with observations. Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca2, choices=1:2, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca2, choices=3:4, pc.biplot=T, cex=c(.5, 1), xlim=c(-3,3), ylim=c(-2.5,2.5))
@
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the remaining principal components distinguish between the cognitive tasks; 
PC2 and PC3 separate  paper folding from visual search and from the lineup and figure classification tasks, while PC4 and PC5 mainly separate lineups from card rotation and figure classification. This separation allows us to compare the tasks which are similar from among the principal components. 
According to Table~\ref{tab:PCArotation4}, the first three principal components account for \Sexpr{round(summary(pca2)$importance[3,3]*100,1)}\% of the variance.

@@

% \comment{Describe performance - need to get data from original experiment.}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}
\caption{Boxplots used to compare the two distributions.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}
\caption{Density plots used to compare the two distributions.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}
\caption{Dotplots used to compare the two distributions.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}
\caption{Histograms used to compare the two distributions.}
\end{figure}
\FloatBarrier
\subsection{Lineup Set 2}
The second lineup section also explored two groups of data, this time comparing boxplots, bee swarm boxplots, boxplots with overlaid jittered data, and violin plots. Participants were much more accurate in this experiment than in the experiment described previously, because of the types of plots compared as well as the underlying data distributions. 
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f94f6965d7}
\caption{Boxplots.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f9e65f0e4}
\caption{Boxplots with jittered points.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f9fc58d73}
\caption{Bee swarm boxplots.}
\end{figure}
\begin{figure}[htbp]
\includegraphics[keepaspectratio=T,width=.9\linewidth]{../../Images/Lineups2/file147f93acbb7e4}
\caption{Violin plots.}
\end{figure}
\FloatBarrier
\subsection{Lineup Set 3}
The final lineup section explored QQ-plots from various model simulations, using reference lines, acceptance bands, and rotation to determine which plots allowed participants to most effectively identify violations of normality. Rotated qq plots showed lower performance because participants could more accurately compare acceptance bands to residuals, and thus could identify that the reference bands were too liberal. As a result, performance was somewhat lower for rotated plots, even though participants could more accurately compare the residuals to the reference bands.
% \comment{Add citations, more explanations, etc.}
